\begin{activity}\label{A:0.6.1}
Power functions and exponential functions appear somewhat similar in their formulas, but behave differently in many ways.  
\ba
	\item Compare the functions $f(x)=x^2$ and $g(x)=2^x$ by graphing both functions in several viewing windows.  Find the points of intersection of the graphs.  Which function grows more rapidly when $x$ is large?
    \item Compare the functions $f(x)=x^{10}$ and $g(x)=2^x$ by graphing both functions in several viewing windows.  Find the points of intersection of the graphs.  Which function grows more rapidly when $x$ is large?
	\item Make a conjecture: As $x\to\infty$, which dominates, $x^n$ or $a^x$ for $a>1$?
% 	\item Suppose you are offered a job that lasts one month.  You have the option of being paid in one of two ways: (1) One million dollars at the end of the month; or (2) One cent on the first day of the month, two cents on the second day, four cents on the third day, and, in general, $2^{n-1}$ cents on the $n^{th}$ day.  Which option should you choose?
% 	\item How much different (shorter or longer) would the work period need to be for your answer to the previous question change?
        \ea

\end{activity}
\begin{smallhint}
You may need to use a graphing tool for this activity.
\end{smallhint}
\begin{bighint}
You may need to use a graphing tool for this activity. To see the full behavior of the
functions you will need to use the zoom and pan tools on your graphing technology.
\end{bighint}
\begin{activitySolution}
   \ba
        \item The functions $f(x) = x^2$ and $g(x) = 2^x$ intersect at three points:
            $(-0.77,0.59)$, $(2,4)$, and $(4,16)$.  After the point $(4,16)$ the
            exponential function $g(x) = 2^x$ grows more rapidly.
        \item The functions $f(x) = x^{10}$ and $g(x) = 2^x$ intersect at three points:
            $(-0.94, 0.52)$, $(1.08, 2.11)$, and another point near $x \approx 60$.  The
            $y$ values on both functions at the third point are very very large, but at
            this point the function $g(x) = 2^x$ dominates the power function $f(x) =
            x^{10}$.
        \item For $a>1$ the exponential function will always dominate the power function.
        \item 
   \ea
\end{activitySolution}

\aftera
