\section{Functions, Slope, and Lines} \label{S:0.1.Functions}


\vspace*{-14 pt}
\framebox{\hspace*{3 pt}
\parbox{6.25 in}{\begin{goals}
\item What is a function and what do we mean by its domain and range?
\item What is the slope of a line? What are linear functions and families of linear
    functions?
\item What are difference equations and the delta notation and why are these useful?
\end{goals}} \hspace*{3 pt}}

% \begin{web}
% \item
%     \href{https://www.youtube.com/watch?v=AfoNnJ038zA&list=PL9bIjQJDwfGuXQHuS5Jkmum_CFILoCZX-&index=92}{Video:
%     How to read a math textbook}
% \item
%     \href{https://www.khanacademy.org/math/algebra/solving-linear-equations-and-inequalities}{Khan
%     Playlist: Linear Equations}
% \item \href{https://www.khanacademy.org/math/algebra/algebra-functions}{Khan Playlist:
%     Functions}
% \item \href{https://www.khanacademy.org/math/algebra2/functions_and_graphs}{Khan Playlist:
%     Functions and Graphs}
% \end{web}

\nin \hrulefill



\subsection*{Introduction}
We begin the study of calculus by reminding the reader of several pre-requisite topics.
The study of calculus depends on a thorough understanding of these topics and it is
imperative that the reader become as familiar as possible with these topics.  In the
present section we remind the reader about the concepts of functions, slope, and lines,
but first, there are a few things that you should do to get your self ready to use this
text.

\input{previews/0.1.PA1}

\subsection*{Functions}
\begin{definition}
    A {\bf function} $f$ defined on a set $A$, is a rule that assigns to each element $x$ in $A$,
exactly one element, denoted $f(x)$, from a set $B$. 

\noindent The set $A$ is called the {\bf domain} of the function $f$. The {\bf range} of $f$ is the set
of values of $f(x)$
as $x$ takes on all the values of $A$.  Another way to state it is that the range of $f$ is the
set of all $y$ such that $y=f(x)$ for some $x$ in $A$.
\end{definition}

It is easy to give many common examples of functions:
\begin{itemize}
    \item The area of a circle $A$ is a function of the radius of the circle:  $A(r)= \pi r^2$.
    \item The amount $M$ in your savings account is a function of the rate of interest the
        bank pays.
    \item Your miles per gallon in your car depends on many things, e.g. the speed at
        which you drive.  
    \item The pressure on a diver is a function of the depth of the diver under water.
\end{itemize}

Probably the most common method for representing a function is with a graph.  If the
domain of function $f$ is set $A$, then the graph of $f$ is the collection of all ordered
pairs of the form $(x,f(x))$ where $x$ comes from the domain $A$.



\input{activities/0.1.Act1}

\bex
Find the domain and range of the functions
\[ f(x) = \sin(x), \quad g(x) = \sqrt{x}, \quad \text{and} \quad h(x) =
    \frac{1}{x}. \]
\eex
For $f(x)=\sin(x)$ we recall that the sine function is defined for every possible value of $x$ but
the output is strictly between $y=-1$ and $y=1$.  Therefore, the domain for $f(x) =
\sin(x)$ is $-\infty < x < \infty$ and the range is $-1 \le y \le 1$.  See the left plot
in Figure \ref{f:0.1ex1}.

For $g(x) = \sqrt{x}$ we recall that the square root of a negative number results in an
imaginary number.  In this text we are interested in real-valued output for functions so
we must omit all of the negative numbers from the domain and hence $0 \le x < \infty$.
For the range we recall that the square root of a number will always be a non-negative
number.  As such, the range is $0 \le y < \infty$.  See the middle plot
in Figure \ref{f:0.1ex1}.

For $h(x) = \frac{1}{x}$ we recall that division by zero is mathematically impossible.
That is the only troublesome point in the domain so $-\infty < x < 0$ or $0 < x < \infty$.
A moment's reflection also reveals that it is impossible to get zero out of the function
$h(x)$ but it is possible to get any other number.  Hence $-\infty < y < 0$ or $0 < y <
\infty$. See the right plot in Figure \ref{f:0.1ex1}. 
\afterex

\begin{figure}
    \begin{center}
        \begin{tikzpicture}[scale=0.65]
            \begin{axis}[axis lines=center, title={$f(x) = \sin(x)$}, domain=-6.28:6.28,
                ymin=-1.1, ymax = 1.1, grid]
                \addplot[smooth, very thick, blue] {sin(deg(x))};
            \end{axis}
        \end{tikzpicture}
        \begin{tikzpicture}[scale=0.65]
            \begin{axis}[axis lines=center, title={$g(x) = \sqrt{x}$}, domain=-1:4,
                ymin=-0.5, ymax = 2.5, xmin=-1, grid]
                \addplot[smooth, very thick, red, samples=100] {sqrt(x)};
            \end{axis}
        \end{tikzpicture}
        \begin{tikzpicture}[scale=0.65]
            \begin{axis}[axis lines=center, title={$h(x) = 1/x$},
                ymin=-2, ymax = 2, grid]
                \addplot[smooth, very thick, domain=-3:-0.1, black] {1/x};
                \addplot[smooth, very thick, domain=0.1:3, black] {1/x};
            \end{axis}
        \end{tikzpicture}
    \end{center}
    \caption{Graphs of the function $f(x) = \sin(x)$, $g(x) = \sqrt{x}$, and $h(x) =
    \frac{1}{x}$.}
    \label{f:0.1ex1}
\end{figure}

\subsection*{Slope and Linear Functions}
In Calculus we are often interested in writing the equation for a linear function.  As
such, we should review the features of linear functions.  Every linear function is
characterized by a constant rate of change; the slope.  The slope of a linear function is
a measure of the ``steepness'' of the line.  We use the symbols $\Delta x$,
$\Delta y$ which
mean respectively the ``change in $x$'' and the ``change in $y$''.  
\begin{definition}
    The {\bf slope}, $m$ of a (non-vertical) linear function $f$ which passes through any
    two points $(x_1,y_1)$, $(x_2,y_2)$ can be found using the formula
    \[ m = \frac{\Delta y}{\Delta x} = \frac{y_2 - y_1}{x_2 - x_1} =
    \frac{f(x_2)-f(x_1)}{x_2-x_1} = \frac{\text{Rise}}{\text{Run}} \]
\end{definition}

\begin{minipage}{0.4\columnwidth}
    Recall that
\begin{itemize}
    \item if the line rises from left to right then the slope is positive,
    \item if the line falls from left to right then the slope is negative,
    \item if the line is horizontal then the slope is zero, and
    \item if the line is vertical then the slope is undefined.
\end{itemize}
\end{minipage}
\begin{minipage}{0.5\columnwidth}
    \begin{center}
        \begin{tikzpicture}[scale=0.8]
            \draw[<->, thick] (-4,0) -- (4,0) node[anchor=west]{$x$};
            \draw[<->, thick] (0,-2) -- (0,4) node[anchor=south]{$y$};
            \draw[<->, very thick, blue] (-4,3) -- (4,-1) node[anchor=west]{Negative
            Slope};
            \draw[<->, very thick, color=red, dashed] (-4,-2) -- (4,3)
            node[anchor=west]{Positive Slope};
            \draw[<->, very thick, color=black, dotted] (-4,2) -- (4,2)
            node[anchor=west]{Zero Slope};
        \end{tikzpicture}
    \end{center}
\end{minipage}\bigskip

Depending on the information given there are several convenient forms of the equation of a
line.  Given the definition of the slope
\[ m = \frac{y_2 - y_1}{x_2 - x_1} \]
and letting $(x,y) = (x_2,y_2)$ be any arbitrary point we get the point-slope form of a
linear function.
\begin{definition}
If the linear function $f$ has slope $m$ and passes through the
point $(x_1,y_1)$, then the {\bf point-slope form of the equation of a line} is given by: 
\[ y-y_1=m(x- x_1).  \]
\end{definition}

An alternate form of a linear function which is probably very familiar to most readers is
the slope-intercept form of a line.
\begin{definition}
If the linear function $f$ has slope $m$ and $y$-intercept $b$, then the
{\bf slope-intercept form of the equation of a line} is given by: 
\[ y=mx + b.  \]
\end{definition}
In a calculus class the point-slope form is often the most useful. The
symbols and geometry used in each of the above definitions are shown in Figure
\ref{fig:0.1.linear_fn}.

\begin{figure}[ht!]
    \centering
    \begin{tikzpicture}
        \draw[color=gray!50] (-4,-2) grid (4,4);
        \draw[<->, thick] (-4,0) -- (4,0) node[anchor=west]{$x$};
        \draw[<->, thick] (0,-2) -- (0,4) node[anchor=south]{$y$};
        \draw[<->, very thick, blue] (-4,3) -- (4,-1);
        \draw[color=black, fill=black] (-3.5,2.75) circle(0.075cm) node[anchor=south
        west]{$(x_1,y_1)$};
        \draw[color=black, fill=black] (-1,1.5) circle(0.075cm) node[anchor=south
        west]{$(x_2,y_2)$};
        \draw[color=black, fill=black] (0,1) circle(0.075cm) node[anchor=south
        west]{$(0,b)$};
        \draw[thick, dashed] (-3.5,2.75) -- (-3.5,1.5) -- (-1,1.5);
        \draw (-3.5,2.25) node[anchor=east]{Rise$=\Delta y$};
        \draw (-2,1.5) node[anchor=north]{Run$=\Delta x$};
    \end{tikzpicture}
    \caption{Anatomy of a linear function.}
    \label{fig:0.1.linear_fn}
\end{figure}


\input{activities/0.1.Act2}

\bex
Write the equation of the line going through the points $(5,7)$ and $(-3,2)$.
\eex
First we calculate the slope 
\[ m = \frac{\Delta y}{\Delta x} = \frac{7 - 2}{5-(-3)} = \frac{5}{8}. \]
Since we have two points and neither is the $y$ intercept of the linear function we choose
to use the point-slope form of the line.  Letting $(x_1,y_1) = (5,7)$ we see that
\[ y - 7 = \frac{5}{8} \left( x-5 \right) \]
is one form of the linear function.  
It is often conventient to solve for $y$ giving us
\[ y = \frac{5}{8} \left( x-5 \right) + 7. \]
Notice that we do not necessarily need to simplify all the way to the slope-intercept form
of the line.  
\afterex

\subsection*{Linear Functions From Data}
A feature of every linear function is that the slope is the same no matter where you are
on the line.  When given a table of data that you suspect might represent a linear
function the slope manifests itself as a constant common difference between successive
$y$-values.  

\bex
Consider the data in the table below.
\begin{center}
    \begin{tabular}[h!]{|c||c|c|c|c|c|}
        \hline
        $x$ & 5 & 6 & 7 & 8 & 9 \\ \hline
        $y$ & 12.2 & 17.5 & 22.8 & 28.1 & 33.4 \\ \hline
    \end{tabular}
\end{center}
Demonstrate that this data is linear and write an equation that fits the data.
\eex
The common differences can be found for each successive $y$-values
\begin{center}
    \begin{tabular}[h!]{|c||c|c|c|c|c|}
        \hline
        $x$ & 5 & 6 & 7 & 8 & 9 \\ \hline
        $y$ & 12.2 & 17.5 & 22.8 & 28.1 & 33.4 \\ \hline
        Common Difference & $\frac{17.5-12.2}{6-5} = 5.3$ & $\frac{22.8-17.5}{7-6} = 5.3$
        & $\frac{28.1-22.8}{8-7} = 5.3$ & $\frac{33.4-28.1}{9-8}=5.3$ & - \\ \hline
    \end{tabular}
\end{center}
The successive differences are clearly the same throughout the data set and the slope for
this data set is $m=5.3$.  Picking any convenient point, say $(5,12.2)$, then allows us to write the
equation of the line as 
\[ y - 12.2 = 5.3(x-5). \]
This could be simplified to point-slope form, but there is typically no need for this
algebraic simplification.
\afterex


\input{activities/0.1.Act3}


\bex
The Old Farmer's Almanac tells us that you can tell the temperature by counting the chirps
of a cricket.  It is a linear function $T=f(C)$ given by $T$ (in degrees Fahrenheit)=\# of
chirps in 15 seconds $+40$.  We can approximate this with the formula 
\[ T =  \frac{C}{4} + 40 \]
where $C$ is the number of chirps/minute and $T$ is in $^\circ F$.
\ba
    \item If the chirp rate is 120 chirps/minute, what is the temperature?
    \item Suppose that crickets will not chirp if the temperature is below $56^\circ F$.
        We can also suppose that crickets will not chirp above $136^\circ F$ since that is
        the highest temperature ever recorded at a weather station.  With these
        parameters, what is the domain of this function?
\ea
\eex
\ba
    \item If $C = 120$ chirps/minute, substitute this into the function $T(C)$ to obtain
        \[ T(120) = \frac{120}{4} + 40 = 30 + 40 = 70^\circ F. \]
    \item To find the domain we need to find the appropriate values of $C$ for the $T(C)$
        function.  Solve $56=C/4+40$ and get $C = 64$.  Solve $136=C/4+40$ and get $C =
        384$.  So the domain of $T(C)$ is $64 \le \text{chirps/minute} \le 384$ or, in
        interval notation, $[64, 136]$. 

\ea
\afterex


\subsection*{Families of Linear Functions}
We noted above that a linear function has the form  $y=f(x)=mx+b$, where $m$ is the slope of
the line, and $b$ is the $y$-intercept.  Since $m$ and $b$ can take on various values, taken
together, they represent a family of functions.  For example, we could fix $b = 2$, and then
draw the graphs of $f(x)=mx+2$ for various values of $m$; for example, $m = -1, -2, 2, 1$.
Doing so would give the functions in the family $f(x)=mx+2$ shown in the left image of Figure
\ref{fig:0.1.fam1}.

Similarly, we could set $m$ to be $2$ and let $b$ take on the values $b=-1, 1, 4, -6$ and
we would get
some examples from the family of functions for $y=f(x)=2x+b$ shown in the right image of Figure
\ref{fig:0.1.fam1}.


From the right image in Figure \ref{fig:0.1.fam1} it should be clear to the reader that
parallel lines have the same slope.  What can you say about the slopes of perpendicular
lines?  Here is the result that we state without proof.
\begin{theorem}\label{thm:test}
If line $\ell_1$ has slope $m_1$ and line $\ell_2$ has slope $m_2$, then 
    \begin{itemize}
        \item lines $\ell_1$ and $\ell_2$ are parallel if the slopes are the same: $m_1 = m_2$,
            and
        \item lines $\ell_1$ and $\ell_2$ are perpendicular if the slopes are opposite
            reciprocals: $m_2 = -\frac{1}{m_1}$.
    \end{itemize}
\end{theorem}

\def\scl{0.8}
\begin{figure}[ht]
    \centering
    \begin{tikzpicture}[scale=\scl]
        \begin{axis}[axis lines=center, xmin=-5, xmax=5, ymin=-10, ymax=10,
            legend style={ at={(axis cs:4.03,5)}, anchor=west }]
            \addplot[smooth, very thick, color=blue] {-1*x+2};
            \addlegendentry{$y=-x+2$};
            \addplot[smooth, very thick, color=red, dashed] {-2*x+2};
            \addlegendentry{$y=-2x+2$};
            \addplot[smooth, very thick, color=cyan, dotted] {1*x+2};
            \addlegendentry{$y=x+2$};
            \addplot[smooth, very thick, color=black] {2*x+2};
            \addlegendentry{$y=2x+2$};
        \end{axis}
    \end{tikzpicture}
    \begin{tikzpicture}[scale=\scl]
        \begin{axis}[axis lines=center, xmin=-5, xmax=5, ymin=-10, ymax=10,
            legend style={ at={(axis cs:4.03,5)}, anchor=west }]
            \addplot[smooth, very thick, color=blue] {2*x-1};
            \addlegendentry{$y=2x-1$};
            \addplot[smooth, very thick, color=red, dashed] {2*x+1};
            \addlegendentry{$y=2x+1$};
            \addplot[smooth, very thick, color=cyan, dotted] {2*x+4};
            \addlegendentry{$y=2x+4$};
            \addplot[smooth, very thick, color=black] {2*x-6};
            \addlegendentry{$y=2x-6$};
        \end{axis}
    \end{tikzpicture}
    \caption{Several members of the family of linear functions $f(x) = mx+2$ (left) and
    the family $f(x) = 2x+b$ (right).}
    \label{fig:0.1.fam1}
\end{figure}

\input{activities/0.1.Act4}


% \subsection*{Equations, Functions, and Expressions}
% To conclude this first section we define three commonly misused mathematical terms.
% \begin{definition}
%     A {\bf mathematical expression} is a combination of numbers, variables, and operations
%     (addition, subtraction, multiplication, roots, etc).  Several examples are:
%     \begin{itemize}
%         \item $2x+3$
%         \item $\sqrt{a^2 + b^2}$
%         \item $\pi r^2$
%     \end{itemize}
% \end{definition}
% 
% \begin{definition}
%     An {\bf equation} is a mathematical statement containing an equal sign where the left
%     and right-hand sides of the equal sign are expressions in the same variable.  Examples
%     are:
%     \begin{itemize}
%         \item $3=5x-2$
%         \item $x^2-2x+5=9$
%         \item $\sqrt{3x-2} = 19$
%     \end{itemize}
% \end{definition}
% 
% \begin{definition}
%     A {\bf function} (as defined before) is a rule that associates a value $x$ in the
%     domain to a value $y$ in the range.  Examples include:
%     \begin{itemize}
%         \item $y=x^2+3$
%         \item $f(x) = 3x-4$
%         \item $f(y) = \sqrt{3y-2}$
%     \end{itemize}
% \end{definition}
% 
% It is easy for students to confuse the meanings of these three definitions, so here are
% some helpful tips:
% \begin{itemize}
%     \item If there is no equal sign then it is an expression.
%     \item It makes sense to substitute a value into an expression, but saying that you're
%         going to ``solve'' an expression is meaningless.
%     \item If it is an equation then it is meaningful to ``solve'' the equation.  It is NOT
%         meaningful to say that you are going to ``solve'' a function or expression.
%     \item A function defines a rule that associates two variables.
%     \item A function has an associated graph whereas expressions and equations do not.
% \end{itemize}
% 
% \input{activities/0.1.Act5}
% 
% % more activites

\begin{summary}
\item A function assigns one $y$ value to each $x$ value.
\item The slope of a linear function can be written as 
    \[ m = \frac{\text{Rise}}{\text{Run}} = \frac{y_2 - y_1}{x_2 - x_2} \]
\item A linear function can be written in the forms
    \[ y = mx + b \quad \text{or} \quad y-y_1 = m(x-x_1) \]
\item When examining linear data, the differences between successive $y$-values reveals
    the slope.
\end{summary}

\nin \hrulefill

% exercises go here
\input{exercises/0.1.FunctionsLinesSlope(Ex)}






\clearpage

